---
title: The ToxCast(TM) Analysis Pipeline(tcpl)<br /> An R Package for Processing and Modeling Chemical Screening Data (Version 3.0)<br />
author: "Center for Computational Toxicology and Exposure, US EPA"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{The ToxCast(TM) Analysis Pipeline(tcpl)<br /> An R Package for Processing and Modeling Chemical Screening Data (Version 3.0)<br />}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
<!-- This CSS script ensures autonumbering of equations-->
<script type="text/x-mathjax-config">
  MathJax.Hub.Config({ TeX: { equationNumbers: {autoNumber: "all"} } });
</script>


<script>
$(document).ready(function() {
  $items = $('div#TOC li');
  $items.each(function(idx) {
    num_ul = $(this).parentsUntil('#TOC').length;
    $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
  });

});
</script>

```{r eval = FALSE, echo = FALSE, message = FALSE, warning = FALSE}

library(htmlTable)
```

# Introduction

This vignette provides an overview of the <font face="CMTT10"> tcpl </font> package, including set up, database structure, and pre-processing. Assay registration, data processing, and data retrieval are topics covered in separate vignettes.

# Overview

The ToxCast pipeline (<font face="CMTT10"> tcpl </font>) is an R package that manages, curve-fits, plots, and stores ToxCast data to populate its linked MySQL database, InvitroDB. The U.S. Environmental Protection Agency (EPA) ToxCast^TM^ program. includes in vitro medium- and high-throughput screening assays for the prioritization and hazard characterization of thousands of chemicals of interest.These assays comprise Tier 2-3 of the new Computational Toxicology Blueprint, and employ automated chemical screening technologies, to evaluate the effects of chemical exposure on living cells and biological macromolecules, such as proteins (Thomas et al., 2019). More information on the ToxCast program can be found at <http://www.epa.gov/chemical-research/toxicity-forecasting>.

This flexible analysis pipeline is capable of efficiently processing and storing large volumes of data. The diverse data, received in heterogeneous formats from numerous vendors, are transformed to a standard computable format and loaded into the <font face="CMTT10"> tcpl </font> database by vendor-specific R scripts. Once data is loaded into the database, ToxCast utilizes generalized processing functions provided in this package to process, normalize, model, qualify, and visualize the data. 

The original <font face="CMTT10"> tcplFit() </font> functions performed basic concentration response curve fitting in past versions of <font face="CMTT10"> tcpl </font>. Processing with <font face="CMTT10"> tcpl_v3 </font> onward encompass the stand-alone <font face="CMTT10"> tcplFit2 </font> package to allow a wider variety of concentration response models.The main set of extensions was to include all of the concentration-response models that are contained in the program BMDExpress. These include exponential and power functions in addition to the original Hill, gain-loss and constant models. Additionally, we wanted to include BMD (Benchmark Dose Modeling) outputs, which is simply defining a Benchmark Response (BMR) level and setting the BMD to the concentration where the curve crosses the BMR level. One final addition was to let the hitcall value be a continuous number ranging from 0 to 1. While developed primarily for ToxCast, the <font face="CMTT10"> tcpl </font> package is written to be generally applicable to the chemical-screening community.

The <font face="CMTT10"> tcpl </font> package includes processing functionality for two screening paradigms: (1) single-concentration screening and (2) multiple-concentration screening. Single-concentration screening consists of testing chemicals at one concentration, often for the purpose of identifying potentially active chemicals to test in the multiple-concentration format. Multiple-concentration screening consists of testing chemicals across a concentration range, such that the modeled activity can give an estimate of potency, efficacy, etc. 

Prior to the pipeline processing provided in this package, all the data must go through pre-processing (level 0). Level 0 pre-processing utilizes dataset-specific R scripts to process the heterogeneous data into a uniform format and to load the uniform data into the <font face="CMTT10"> tcpl </font> database. Level 0 pre-processing is outside the scope of this package, but can be done for virtually any high-throughput or high-content chemical screening effort, provided the resulting data includes the minimum required information. 

In addition to storing the data, the <font face="CMTT10"> tcpl </font> database stores every processing and analysis decision at the assay component or assay endpoint level to facilitate transparency and reproducibility. For the illustrative purposes of this vignette and others, we have included a CSV version of the <font face="CMTT10"> tcpl </font> database containing a small subset of data from the ToxCast program. tcplLite, which allowed the user to upload a flat file in ToxCast database format and process the data using the tcpl analysis protocols, is no longer supported by <font face="CMTT10"> tcpl </font>. <font face="CMTT10"> tcplFit2 </font>, however, can be used for stand-alone applications, available at <https://cran.r-project.org/web/packages/tcplfit2/index.html>. Using <font face="CMTT10"> tcpl </font>, the user can upload, process, and retrieve data by connecting to a MySQL database. The package includes a SQL file to initialize a blank schema on the user's server of choice. Additionally, past versions of the ToxCast database, containing all the publicly available ToxCast data, are available for download at: <https://www.epa.gov/chemical-research/exploring-toxcast-data-downloadable-data>.

# Package Settings


First, it is highly recommended for users to utilize the <font face="CMTT10"> data.table </font> package. The <font face="CMTT10"> tcpl </font> package utilizes the <font face="CMTT10"> data.table </font> package for all data frame-like objects. 
```{r eval=TRUE, message=FALSE}
library(data.table)
library(tcpl)
## Store the path for the tcpl directory for loading data
pkg_dir <- system.file(package = "tcpl")

```
Every time the package is loaded in a new R session, a message similar to the following will print showing the default package settings:

```{r eval = FALSE}
tcpl (v1.3) loaded with the following settings:
  TCPL_DB:    C:/Users/user/R-3.4.4/library/tcpl/csv
  TCPL_USER:  NA
  TCPL_HOST:  NA
  TCPL_DRVR:  NA
Default settings stored in TCPL.conf. See ?tcplConf for more information.
```

The package consists of five settings: 

1. <font face="CMTT10"> $TCPL_DB </font> points to the <font face="CMTT10"> tcpl </font> database (either the path to the CSV directory, as in the given example above, or the name of the MySQL database),
2. <font face="CMTT10"> $TCPL_USER </font> stores the username for accessing the database,
3. <font face="CMTT10"> $TCPL_PASS </font> stores the password for accessing the database,
4. <font face="CMTT10"> $TCPL_HOST </font> points to the MySQL server host, and 
5. <font face="CMTT10"> $TCPL_DRVR </font> indicates which database driver is used ("MySQL"). tcplLite is no longer supported and it is recommended to use <font face="CMTT10"> tcplFit2 </font> package for stand-alone applications.

Refer to <font face="CMTT10"> ?tcplConf </font> for more information. At any time, users can check the settings using <font face="CMTT10"> tcplConfList() </font>. An example of database settings using tcpl would be as follows:

```{r eval = FALSE}
tcplConf(db   = "invitrodb",
         user = "username", 
         pass = "password", 
         host = "localhost",
         drvr = "MySQL")
```


<font face="CMTT10"> tcplConfList </font> will list connection information. Note, <font face="CMTT10"> tcplSetOpts </font> will only make changes to the parameters given. The package is always loaded with the settings stored in the TCPL.config file located within the package directory. The user can edit the file, such that the package loads with the desired settings, rather than having to call the <font face="CMTT10"> tcplSetOpts </font> function every time. The TCPL.config file has to be edited whenever the package is updated or re-installed.

# Database Structure

The following contains reference tables that describe the structure and fields found in the <font face="CMTT10"> tcpl </font> populated database. The first sections describe the data-containing tables, followed by sections describing the additional annotation tables.

In general, the single-concentration data and accompanying methods are found in the "sc#" tables, where the number indicates the processing level. Likewise, the multiple-concentration data and accompanying methods are found in the "mc#" tables. Each processing level that has accompanying methods will also have tables with the "_methods" and "_id" naming scheme. For example, the database contains the following tables: "mc5" storing the data from multiple-concentration level 5 processing, "mc5_methods" storing the available level 5 methods, and "mc5_aeid" storing the method assignments for level 5. Note, the table storing the method assignments for level 2 multiple-concentration processing is called "mc2_acid", because MC2 methods are assigned by assay component ID.

There are two additional tables, "sc2\_agg" and "mc4\_agg," that link the data in tables "sc2" and "mc4" to the data in tables "sc1" and "mc3," respectively. This is necessary because each entry in the database before SC2 and MC4 processing represents a single value; subsequent entries represent summary/modeled values that encompass many values. To know what values were used in calculating the summary/modeled values, the user must use the "\_agg" look-up tables.

<font face="CMTT10"> tcpl_v3 </font> includes new "mc4_param" and "mc5_param" tables to capture model fit parameters from <font face="CMTT10"> tcpfit2 </font>. mc4_param includes parameters for all models, whereas mc5_param includes model parameter from selected winning (hit) model.

Each of the methods tables have fields analogous to $\mathit{mc5\_mthd\_id}$, $\mathit{mc5\_mthd}$, and $\mathit{desc}$. These fields represent the unique key for the method, the abbreviated method name (used to call the method from the corresponding <font face="CMTT10"> mc5\_mthds </font> function), and a brief description of the method, respectively. The method assignment tables will have fields analogous to $\mathit{mc5\_mthd\_id}$ matching the method ID from the methods tables, an assay component or assay endpoint ID, and possibly an $\mathit{exec\_ordr}$ field indicating the order in which to execute the methods. The method and method assignment tables will not be listed in the tables below to reduce redundancy. 

Many of the tables also include the $\mathit{created\_date}$, $\mathit{modified\_date}$, and $\mathit{modified\_by}$ fields that store helpful information  for tracking changes to the data. These fields will not be discussed further or included in the tables below.

Many of the tables specific to the assay annotation are not utilized by the <font face="CMTT10"> tcpl </font> package. The full complexity of the assay annotation used by the ToxCast program is beyond the scope of this vignette and the  <font face="CMTT10"> tcpl </font> package. Additionally, assay description documents for ToxCast assays can be found at: <https://www.epa.gov/chemical-research/exploring-toxcast-data-downloadable-data>.

## Single-concentration Data-containing Tables
<!-- Table 1-->
```{r warning = FALSE, echo = FALSE}
Field <- c("s0id ", "acid", "spid", "apid", "rowi", "coli", "wllt", "wllq", "conc", "rval", "srcf")
Description <- c("Level 0 ID",
                 "Assay component ID",
                 "Sample ID",
                 "Assay plate ID",
                 "Assay plate row index",
                 "Assay plate column index",
                 "Well type&dagger;",
                 "1 if the well quality was good, else 0&ddagger;",
                 "Concentration is micromolar",
                 "Raw assay component value/readout from vendor",
                 "Filename of the source file containing the data"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 1: Fields in sc0 table.",
        tfoot="&dagger;Information about the different well types is available in Appendix B.")

```

<!-- Table 2-->
```{r warning = FALSE, echo = FALSE}
Field <- c("s1id ", "s0id", "acid", "aeid", "logc", "bval", "pval", "resp")
Description <- c("Level 1 ID",
                 "Level 0 ID",
                 "Assay component ID",
                 "Assay component endpoint ID",
                 "Log base 10 concentration",
                 "Baseline value",
                 "Positive control value",
                 "Normalized response value"
              
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 2: Fields in sc1 table."
)

```

<!-- Table 3-->
```{r warning = FALSE, echo = FALSE}
Field <- c("aeid ", "s0id", "s1id", "s2id")
Description <- c("Assay component endpoint ID",
                 "Level 0 ID",
                 "Level 1 ID",
                 "Level 2 ID"
              
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 3: Fields in sc2_agg table."
)

```

<!-- Table 4-->
```{r warning = FALSE, echo = FALSE}
Field <- c("s2id ", "aeid", "spid", "bmad", "max_med", "hitc", "coff", "tmpi")
Description <- c("Level 2 ID",
                 "Assay component endpoint ID",
                 "Sample ID",
                 "Baseline median absolute deviation",
                 "Maximum median response value",
                 "Hit-/activity-call, 1 if active, 0 if inactive",
                 "Efficacy cutoff value",
                 "Ignore, temporary index used for uploading purposes"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 4: Fields in sc2 table."
)

```

<!-- Table 5-->
```{r warning = FALSE, echo = FALSE}
Field <- c("s2id", "chid_rep")

Description <- c("Level 2 ID",
                 "1 if representative sample, else 0"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 5: Fields in sc2_chid." 
     
)

```


## Multiple-concentration Data-containing Tables
<!-- Table 6-->
The "mc0" table, other than containing $\mathit{m0id}$ rather than $\mathit{s0id}$, is identical to the "sc0" described in the section above.
```{r warning = FALSE, echo = FALSE}
Field <- c("m1id", "m0id", "acid", "cndx", "repi")
Description <- c("Level 1 ID",
                 "Level 0 ID",
                 "Assay component ID",
                 "Concentration index",
                 "Replicate index"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 6: Fields in mc1 table."
)

```

<!-- Table 7-->
```{r warning = FALSE, echo = FALSE}
Field <- c("m2id", "m0id", "acid", "m1id", "cval")
Description <- c("Level 2 ID",
                 "Level 0 ID",
                 "Assay component ID",
                 "Level 1 ID",
                 "Corrected value"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 7: Fields in mc2 table."
)

```
<!-- Table 8-->
```{r warning = FALSE, echo = FALSE}
Field <- c("m3id", "aeid", "m0id", "acid", "m1id", "m2id", "bval", "pval", "logc", "resp")
Description <- c("Level 3 ID",
                 "Assay endpoint ID",
                 "Level 0 ID",
                 "Assay component ID",
                 "Level 1 ID",
                 "Level 2 ID",
                 "Baseline value",
                 "Positive control value",
                 "Log base 10 concentration",
                 "Normalized response value"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 8: Fields in mc3 table."
)

```
<!-- Table 9-->
```{r warning = FALSE, echo = FALSE}
Field <- c("aeid", "m0id", "m1id", "m2id", "m3id", "m4id")
Description <- c(
   "Assay endpoint ID","Level 0 ID",
                 "Level 1 ID",
                 "Level 2 ID",
                 "Level 3 ID",
                 "Level 4 ID"
                 
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 9: Fields in mc4_agg table."
)

```
<!-- Table 10-->
```{r warning = FALSE, echo = FALSE}

Field <- c("m4id", "aeid", "spid", "bmad", "resp_max", "resp_min", "max_mean", "max_mean_conc", "max_med", "max_med_conc", "logc_max", "logc_min", 
           "nconc", "npts", "nrep", "nmed_gtbl", "tmpi")


Description <- c("Level 4 ID",
                 "Assay endpoint ID",
                 "Sample ID",
                 "Baseline median absolute deviation",
                 "Maximum response value",
                 "Minimum response value",
                 "Maximum mean response value",
                 "Log concentration at *max_mean*",
                 "Maximum median response value",
                 "Log concentration at *max_med*",
                 "Maximum log concentration tested",
                 "Minimum log concentration tested",
                 "Number of concentrations tested ",
                 "Number of points in the concentration series",
                 "Number of replicates in the concentration series",
                 "Number of median values greater than *3bmad*",
                 "Ignore, temporary index used for uploading purposes"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 10: Fields in mc4 table."
)

```


<!-- Table 11-->
```{r warning = FALSE, echo = FALSE}
Field <- c("m4id", "aeid", "model", "model_param", "model_val")

Description <- c("Level 4 ID",
                 "Assay endpoint ID",
                 "Model that was fit",
                 "Key for the parameter that was fit with the corresponding model",
                 "Value for the associated key in the corresponding model"
                )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 11: Fields in mc4_param table."
)

```

<!-- Table 12-->
```{r warning = FALSE, echo = FALSE}

Field <- c("m5id", "m4id", "aeid", "modl", "hitc", "fitc", "coff", "actp", "model_type")


Description <- c("Level 5 ID",
                 "Level 4 ID",
                 "Assay endpoint ID",
                 "Winning model",
                 "Hit-/activity-call, generally a continuous value from 0 to 1 if using tcplfit2 fitting" ,
                 "Fit category",
                 "Efficacy cutoff value",
                "Activity probability (1 - *const_prob* not used with tcplfit2)",
                "Model type placeholder for use when number of fitting methodologies increases"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 12: Fields in mc5 table."
)

```


```{r warning = FALSE, echo = FALSE}
Field <- c("m5id", "aeid", "hit_param", "hit_val")

Description <- c("Level 5 ID",
                 "Assay endpoint ID",
                 "Key for the parameter that was fit with winning model",
                 "Value for the associated key in the winning model"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 13: Fields in mc5_param table."
)

```

<!-- Table 14-->
```{r warning = FALSE, echo = FALSE}
Field <- c("m5id", "chid_rep")

Description <- c("Level 5 ID",
                 "1 if representative sample, else 0" )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 14: Fields in mc5_chid." )

```

## Assay and Auxiliary Annotation Tables
 
The definition of an "assay" is, for the purposes of this package, broken into:

**assay_source**: the vendor/origination of the data 

**assay**: the procedure to generate the component data  

**assay_component**: the raw data readout(s)  

**assay_component_endpoint**: the normalized component data  

Each assay element is represented by a separate table in the <font face="CMTT10"> tcpl </font> database. In general, we refer to an "assay_component_endpoint" as an "assay endpoint." As we move down the hierarchy, each additional layer may have a one-to-many relationship with the previous layer. For example, an assay source can include multiple assays. Given bidirectional fitting in <font face="CMTT10"> tcpl_v3 </font>, a single assay endpoint will derive from each assay component.

All processing occurs by assay component or assay endpoint, depending on the processing type (single-concentration or multiple-concentration) and level. No data are stored at the assay or assay source level. The “assay” and “assay_source” tables store annotations to help in the processing and down-stream understanding of the data.

Throughout the package, the levels of assay hierarchy are defined and referenced by their primary keys (IDs) in the <font face="CMTT10"> tcpl </font> database: $\mathit{asid}$ (assay source ID), $\mathit{aid}$ (assay ID), $\mathit{acid}$ (assay component ID), and $\mathit{aeid}$ (assay endpoint ID). In addition, the package abbreviates the fields for the assay hierarchy names. The abbreviations mirror the abbreviations for the IDs with "nm" in place of "id" in the abbreviations, e.g. assay\_component\_name is abbreviated $\mathit{acnm}$.

A full description of the assay annotation is beyond the scope of this vignette. The fields pertinent to the <font face="CMTT10"> tcpl </font> package are listed in the tables below.

<!-- Table 15-->
```{r warning = FALSE, echo = FALSE}
Field <- c("assay", "assay_component", "assay_component_endpoint", "assay_component_map", "assay_reagent**", "assay_reference**", "assay_source", "chemical", "chemical_library", "citations**", "gene**", "intended_target**", "organism**", "sample")

Description <- c("Assay-level annotation",
                 "Assay component-level annotation",
                 "Assay endpoint-level annotation",
                 "Assay component source names and their corresponding assay component ids",
                 "Assay reagent information",
                 "Map of citations to assay",
                 "Assay source-level annotation",
                 "List of chemicals and associated identifiers",
                 "Map of chemicals to different chemical libraries",
                 "List of citations",
                 "Gene identifiers and descriptions",
                 "Intended assay target at the assay endpoint level",
                 "Organism identifiers and descriptions",
                 "Sample ID information and chemical ID mapping")

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        caption="Table 15: List of annotation tables.",
        tfoot = "** indicates tables not currently used by the *tcpl* package",
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em '
)

```

<!-- Table 16-->
```{r warning = FALSE, echo = FALSE}
Field <- c("aid", "asid&dagger;", "assay_name&dagger;", "assay_desc", "timepoint_hr", 
            "organism_id", "organism",'tissue',"cell_format",
            'cell_free_component_source',
            'cell_short_name', 
            'cell_growth_mode',
            "assay_footprint&dagger;", 
            "assay_format_type" ,
            "assay_format_type_sub" ,
            "content_readout_type",  
            "dilution_solvent" , 
            "dilution_solvent_percent_max")

Description <- c("Assay ID",
                 "Assay source ID",
                 "Assay name (abbreviated \"anm\" within the package)",
                 "Assay description",
                 "Treatment duration in hours",
                 "NCBI taxonomic identifier, available here <https://www.ncbi.nlm.nih.gov/taxonomy>",
                "Organism of origin",
                "Tissue of origin", "Description of cell format",
                "Description of source for targeted cell-free components",
                "Abbreviation of cell line",
                "Cell growth modality", 
                "Microtiter plate size",
                "General description of assay format",
                "Specific description of assay format" ,
                "Description of well characteristics being measured", 
                "Solvent used in sample dilution",
                "Maximum percent of dilution solvent used, from 0 to 1.")

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        caption="Table 16: Fields in assay.",
        tfoot = "&dagger; Required fields for registration",
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em '
)

```

<!-- Table 17-->
```{r warning = FALSE, echo = FALSE}
Field <- c("acid", "aid&dagger;", "assay_component_name&dagger;", "assay_component_desc", "assay_component_target_desc", "parameter_readout_type","assay_design_type", "assay_design_type_sub", "biological_process_target", "detection_technology_type", "detection_technology_type_sub", "detection_technology", "key_assay_reagent_type", "key_assay_reagent", "technological_target_type", "technological_target_type_sub")

Description <- c("Assay component ID",
                 "Assay ID",
                 "Assay component name (abbreviated \"acnm\" within the package)",
                 "Assay component description", 
                 "Assay component target description. Generally includes information about mechanism of action with assay target, how disruption is detected, or significance of target disruption.",
                 "Description of parameters measured", 
                "General description of the biological or physical process is translated into a detectable signal by assay mechanism",
                "Specific description of method through which a biological or physical process is translated into a detectable signal measured",
                "General biological process being chemically disrupted",
                "General description of assay platform or detection signals measured",
                "Description of signals measured in assay platform",
                "Specific description of assay platform used",
                "Type of critical reactant being measured",
                "Critical reactant measured",
                "General description of technological target measured in assay platform",
                "Specific description of technological target measured in assay platform"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        tfoot = "&dagger; Required fields for registration",
        caption="Table 17: Fields in assay_component."
)

```

<!-- Table 18-->
```{r warning = FALSE, echo = FALSE}
Field <- c("asid&dagger;", "assay_source_name&dagger;", "assay_source_long_name", "assay_source_desc")

Description <- c("Assay source ID",
                 "Assay source name (typically an abbreviation of the assay_source_long_name, abbreviated \"asnm\" within the package)",
                 "Full assay source name", 
                 "Assay source description"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        tfoot = "&dagger; Required fields for registration",
        caption="Table 18: Fields in assay_source."
)

```

<!-- Table 19-->
```{r warning = FALSE, echo = FALSE}
Field <- c("aeid", "acid", "assay_component_endpoint_name", "assay_component_endpoint_desc", "assay_function_type", "normalized_data_type&dagger;", "burst_assay&dagger;", "key_positive_control", "signal_direction", "intended_target_type", "intended_target_type_sub", "intended_target_family", "intended_target_family_sub", "fit_all&dagger;", "cell_viability_assay")
           
Description <- c("Assay component endpoint ID",
                 "Assay component ID",
                 "Assay component endpoint name (abbreviated \"aenm\" within the package)", 
                 "Assay component endpoint description",
                 "Description of targeted mechanism and the purpose of the analyzed readout in relation to others from the same assay",
                 "Normalization approach for which the data is displayed",
                 "Indicator if endpoint is included in the burst distribution (1) or not (0); Burst phenomenon can describe confounding activity, such as cytotoxicity due to non-specific activation of many targets at certain concentrations", 
                 "Tested chemical sample expected to produce activity; Used to assess assay validity",
                 "Directionality of raw data signals from assay (gain or loss); Defines analysis direction",
                "General group of intended targets measured",
                "Specific subgroup of intended targets measured", 
                "Family of intended target measured; Populated on ToxCast chemical activity plot within CompTox dashboard",
                "Specific subfamily of intended target measured",
                "Indicator if all results should be fit, regardless of whether max_med surpasses 3bmad cutoff (1) or not (0)",
                "Indicator of the impact of cytotoxicity in confounding (1) or no cytotoxic impact (0)" )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 19: Fields in assay_component_endpoint.",
        tfoot = "&dagger; Required fields for registration"
)

```

<!-- Table 20-->
```{r warning = FALSE, echo = FALSE}
Field <- c("acid", "acsn")

Description <- c("Assay component ID",
                 "Assay component source name"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 20: Fields in assay_component_map table."
)

```

<!-- Table 21-->
```{r warning = FALSE, echo = FALSE}
Field <- c("chid", "casn", "chnm", "dsstox_substance_id")

Description <- c("Chemical ID&dagger;",
                 "CAS Registry Number",
                 "Chemical name",
                 "Unique identifier from U.S. EPA Distributed Structure-Searchable Toxicity (DSSTox) Database"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 21: Fields in chemical table." ,
        tfoot = "&dagger; this is the DSSTox GSID within the ToxCast data, but can be any integer and will be auto-generated (if not explicitly defined) for newly registered
chemicals"
)

```

<!-- Table 24-->
```{r warning = FALSE, echo = FALSE}
Field <- c("chid", "clib")

Description <- c("Chemical ID",
                 "Chemical library"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 22: Fields in chemical_library table." 
     
)

```

<!-- Table 23-->
```{r warning = FALSE, echo = FALSE}
Field <- c("spid", "chid", "stkc", "stkc_unit", "tested_conc_unit")

Description <- c("Sample ID",
                 "Chemical ID",
                 "Stock concentration" ,
                 "Stock concentration unit",
                 "The concentration unit for the concentration values in the data-containing tables"
                 )

output <- 
  data.frame(Field, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 23: Fields in sample table." 
)

```
The stock concentration fields in the "sample" table allow the user to track the original concentration when the neat sample is solubilized in vehicle before any serial dilutions for testing purposes. The US EPA's ChemTrack application and database supports chemical procurement and sample management for ToxCast in vitro screening efforts.

# Level 0 Pre-processing

Level 0 pre-processing can be done on virtually any high-throughput/high-content screening application. In the ToxCast program, level 0 processing is done in R by vendor/dataset-specific scripts. The individual R scripts act as the "laboratory notebook" for the data, with all pre-processing decisions clearly commented and explained. 

Level 0 pre-processing has to reformat the raw data into the standard format for the pipeline, and also can make manual changes to the data. All manual changes to the data should be very well documented with justification. Common examples of manual changes include fixing a sample ID typo, or changing well quality value(s) to 0 after finding obvious problems like a plate row/column missing an assay reagent. 

Each row in the level 0 pre-processing data represents one well-assay component combination, containing 11 fields. The only field in level 0 pre-processing not stored at level 0 is the assay component source name ($\mathit{acsn}$). The assay component source name should be some concatenation of data from the assay source file that identifies the unique assay components. When the data are loaded into the database, the assay component source name is mapped to assay component ID through the assay\_component\_map table in the <font face="CMTT10"> tcpl </font> database. Assay components can have multiple assay component source names, but each assay component source name can only map to a single assay component. 

<!-- Table 24-->
```{r warning = FALSE, echo = FALSE}
Field <- c("acsn", "spid", "cpid", "apid", "rowi", 
           "coli", "wllt", "wllq", "conc", "rval", "srcf")

Description <- c("Assay component source name",
                 "Sample ID",
                 "Chemical plate ID" ,
                 "Assay plate ID",
                 "Assay plate row index, as an integer",
                 "Assay plate column index, as an integer",
                 "Well type",
                 "1 if the well quality was acceptable, else 0",
                 "Concentration in micromolar",
                 "Raw assay component value or readout from vendor",
                 "Filename of the source file containing the data"
                 )
`N/A` <- c("No", "No", "Yes", "Yes","Yes","Yes", "No", "No", "No&dagger;", "Yes&ddagger;", "No")

output <- 
  data.frame(Field, Description, `N/A`)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        
          caption="Table 24: Required fields in level 0 pre-processing." ,
          tfoot = "The N/A column indicates whether the field can be N/A in the pre-processed data. 
 &dagger; In past versions of tcpl, there were some exceptions where concentrations could be N/A. For tcpl_v3, it is best practice to remove N/A values since N/A values may result in processing errors. 
 &ddagger;If the raw value is N/A, well quality must be 0."
)

```

The well type field is used in the processing to differentiate controls from test compounds in numerous applications, including normalization and definition of the assay noise level. Currently, the <font face="CMTT10"> tcpl </font> package includes the eight well types in Table 29. Package users are encouraged to suggest new well types and methods to better accommodate their data. 

<!-- Table 25-->
```{r warning = FALSE, echo = FALSE}
`Well Type` <- c("t", "c", "p", "n", "m",  "o", "b", "v")

Description <- c("Test compound",
                 "Gain-of-signal control in multiple concentrations",
                 "Gain-of-signal control in single concentration" ,
                 "Neutral/negative control",
                 "Loss-of-signal control in multiple concentrations",
                 "Loss-of-signal control in single concentration",
                 "Blank well",
                 "Viability control"
                 )


output <- 
  data.frame(`Well Type`, Description)

library(htmlTable)
htmlTable(output,
        align = 'l',
        align.header = 'l',
        rnames = FALSE  ,
        css.cell =  ' padding-bottom: 5px;  vertical-align:top; padding-right: 10px;min-width: 5em ',
        caption="Table 25: Well types" 
)

```

The final step in level 0 pre-processing is loading the data into the <font face="CMTT10"> tcpl </font> database. The  <font face="CMTT10"> tcpl </font> package includes the <font face="CMTT10"> tcplWriteLvl0 </font> function to load data into the database. The <font face="CMTT10"> tcplWriteLvl0</font> function maps the assay component source name to the appropriate assay component ID, checks each field for the correct class, and checks the database for the sample IDs with well type "t." Each test compound sample ID must be included in the <font face="CMTT10"> tcpl</font> database before loading data. The <font face="CMTT10"> tcplWriteLvl0</font> also checks each test compound for concentration values.

# Appendix

## A: Cytotoxicity Distribution

Recognizing the susbtantial impact of cytotoxicity in confounding high-throughput and high-content screening results, the <font face="CMTT10"> tcpl</font>  package includes methodology for defining chemical-specific cytotoxicity estimates. Our observations based on ToxCast data suggest a complex, and not-yet fully understood cellular biology that includes non-specific activation of many targets as cells approach death. For example, a chemical may induce activity in an estrogen-related assay, but if that chemical also causes activity in hundreds of other assays at or around the same concentration as cytotoxicity, should the chemical be called an estrogen agonist? The <font face="CMTT10"> tcplCytpPt</font>  function provides an estimate of chemical-specific cytotoxicity points to provide some context to the "burst" phenomenon.

The cytotoxicity point is simply the median AC$_{50}$ for a set of assay endpoints, either given by the user or defined within the <font face="CMTT10"> tcpl</font>  database. By default, the <font face="CMTT10"> tcplCytoPt</font>  function uses the assay endpoints listed in the $\mathit{burst\_assay}$ field of the "assay\_component\_endpoint" table, where 1 indicates including the assay endpoint in the calculation. The "burst" assay endpoints can be indentified by running  <font face="CMTT10"> tcplLoadAeid(fld = "burst\_assay", val = 1) </font>.

In addition to the cytotoxicity point, <font face="CMTT10">tcplCytoPt </font> provides two additional estimates: (1) the MAD of the AC$_{50}$ ($\mathit{modl\_ga}$) values used to calculate the cytotoxicity point, and (2) the global MAD. Note, only active assay endpoints (where the hit call, $\mathit{hitc}$, equals $1$) are included in the calculations. Once the burst distribution (cytotoxicity point and MAD) is defined for each chemical, the global burst MAD is defined as the median of the MAD values. Not every chemcial may be tested in every "burst" assay, so the user can determine the minimum number of tested assays as a condition for the MAD value for a particular chemical to be included in the global MAD calculation. By default, if "aeid" is the vector of assay endpoints used in the calculation, <font face="CMTT10">tcplCytoPt </font> requires the chemical to be tested in at least <font face="CMTT10">floor(0.8 * length(aeid)) </font> assay endpoints to be included in the calculation. The user can specify to include all calculated MAD values (note, there must be at least two active assay endpoints to calculate the MAD) by setting 'min.test' to <font face="CMTT10">FALSE </font>. The 'min.test' parameter also accepts a number, allowing the user to explicitly set the requirement.

The global MAD gives an estimate of overall cytotoxicity window, and allows for a cytotoxicity distrubtion to be determined for chemicals with less than two active "burst" assay endpoints. The cytotoxicity point for chemicals with less than two active "burst" endpoints is set to the value given to the 'default.pt' parameter. By default, the <font face="CMTT10">tcplCytoPt </font> assigns 'default.pt' to 3.^[$10^3 = 1000$, therefore, when using micromolar units, $3$ is equivalent to $1$ millimolar. $1$ millimolar was chosen as an arbitrary high concentration (outside the testing range for ToxCast data), based on the principle that all compounds are toxic if given in high enough concentration.]

## B: Build Variable Matrices

The  <font face="CMTT10">tcplVarMat </font> function creates chemical-by-assay matrices for the level 4 and level 5 data. When multiple sample-assay series exist for one chemical, a single series is selected by the <font face="CMTT10"> tcplSubsetChid </font> function. See <font face="CMTT10">?tcplSubsetChid </font> for more information. <br/>

 1. "modl\_ga" -- The $\log_{10}\mathit{AC_{50}}$ (in the gain direction) for the winning model. <br/>
 2. "hitc" -- The hit call for the winning model. <br/>
 3. "m4id" -- The m4id, listing the concentration series selected by  <font face="CMTT10">tcplSubsetChid </font>. <br\>
 4. "zscore" -- The z-score (described below). <br\>
 5.  "tested\_sc" -- $1$ or $0$, $1$ indicating the chemical/assay pair was tested in the single-concentration format. <br\>
 6. "tested\_mc" -- $1$ or $0$, $1$ indicating the chemical/assay pair was tested in the multiple-concentration format. <br\>
 7. "ac50" -- a modified AC$_{50}$ table (in non-log units) where assay/chemical pairs that were not tested, or tested and had a hit call of $0$ or $-1$ have the value $1e6$. <br\>
 8. "neglogac50" -- $-\log_{10}\frac{\mathit{AC_{50}}}{1e6}$ where assay/chemical pairs that were not tested, or tested and had a hit call of $0$ or $-1$ have the value $0$. <br\>

The z-score calculation is based on the output from <font face="CMTT10"> tcplCytoPt </font> (Appendix C), and is calculated for each AC$_{50}$ value as follows:
$$
\mathit{z-score} = -\frac{\mathit{modl\_ga} - \mathit{cyto\_pt}}{\mathit{global\_mad}}\mathrm{,}
$$
Note: the burst z-score values are multiplied by -1 to make values that are more potent relative to the burst distribution a higher positive z-score.

In addition, additional matrices can be defined by the 'add.vars' parameter in the <font face="CMTT10"> tcplCytoPt </font>  function. The 'add.vars' function will take any level 4 or level 5 field and create the respective matrix.


